---
title: "The Curious Case of a Scale Parameter"
author: "Jyotishka Datta"
date: "July 9, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(cache = TRUE)
```

## A Toy Example 

We consider a 2-dimensional linear regression $Y = X \beta + \epsilon$ where $\beta = (20,20)$ and $X$ is an identity matrix. 
```{r, eval = TRUE}
set.seed(123)
X = matrix(c(1,0,0,1),ncol=2,byrow=T)
beta = c(20,20)
(y = X%*%beta + rnorm(2))
```

We will use two different methods: horseshoe using the `horseshoe` R package and a Bayesian version of scaled / square-root Lasso. 

### Horseshoe results 

```{r}
library(horseshoe)
res <- horseshoe(y, X, method.tau = "truncatedCauchy",
                 method.sigma = "Jeffreys",
                 burn = 1000, nmc = 5000, alpha = 0.05)
(postmean_HS <- res$BetaHat)
(postmedian_HS <- res$BetaMedian)
```

The posterior mean and median estimates are shrunk away from the true mean value, 20. The posterior mode is even more wrong. 

```{r}
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
(map_HS <- apply(res$BetaSamples,1,Mode))
```

This is unexpected given the tail-robustness property of horseshoe prior. The contour plot also shows that the density is concentrated around zero. 

```{r}
theta.smpls.hs <- t(res$BetaSamples)
colnames(theta.smpls.hs) = c("theta1", "theta2")

library(ggplot2)
hs.2d.density<- ggplot(as.data.frame(theta.smpls.hs), aes(theta1, theta2)) +
  geom_density2d()+stat_density2d(aes(fill = ..level..), geom="polygon")+
  xlab(expression(theta[1]))+ylab(expression(theta[2]))+ggtitle("Horseshoe Regression")
print(hs.2d.density)
```

### Bayesian Square Root Lasso Results 

```{r}
setwd("C:/Users/Jyotishka/OneDrive/Documents/R/sqlasso")
source("bayes.sqrt.lasso.R")
res2 <- bayes.sqrt.lasso(Y = y,method.tau ="gamma", r = 1, delta = 2, burn = 1000, nmc = 5000, verbose=FALSE)

(postmean_SQL <- res2$ThetaHat)
```

The mode is close to the true value: 
```{r}
(map_SQL <- apply(res2$ThetaSave,2,Mode))
```

Now, the contour plot: 
```{r}
theta.smpls.sql <- res2$ThetaSave
colnames(theta.smpls.sql) = c("theta1", "theta2")


sql.2d.density<- ggplot(as.data.frame(theta.smpls.sql), aes(theta1, theta2)) +
  geom_density2d()+stat_density2d(aes(fill = ..level..), geom="polygon")+
  xlab(expression(theta[1]))+ylab(expression(theta[2]))+ggtitle("Bayesian Square-root Lasso")
print(sql.2d.density)
```

### Not $\tau$, but $\sigma$

The problem with horseshoe prior doesn't go away if one uses a half-Cauchy (non-truncated) prior on $\tau$ or use the Empirical Bayes method. 

```{r}
res <- horseshoe(y, X, method.tau = "fixed", method.sigma = "Jeffreys",
                 burn = 1000, nmc = 5000, alpha = 0.05)
(postmean_HS <- res$BetaHat)
(postmedian_HS <- res$BetaMedian)
(map_HS <- apply(res$BetaSamples,1,Mode))
```

However, if we use the Empirical Bayes method for estimating $\sigma^2$ instead of using the Jeffrey's prior on $\sigma^2$, the sitaution improves: 
```{r}
res3 <- horseshoe(y, X, method.tau = "fixed", method.sigma = "fixed",
                 burn = 1000, nmc = 5000, alpha = 0.05)
(postmean_HS <- res3$BetaHat)
(postmedian_HS <- res3$BetaMedian)
(map_HS <- apply(res3$BetaSamples,1,Mode))
```

The contour plot looks correct now. 
```{r, echo = FALSE}
theta.smpls.hs <- t(res3$BetaSamples)
colnames(theta.smpls.hs) = c("theta1", "theta2")

library(ggplot2)
hs.2d.density<- ggplot(as.data.frame(theta.smpls.hs), aes(theta1, theta2)) +
  geom_density2d()+stat_density2d(aes(fill = ..level..), geom="polygon")+
  xlab(expression(theta[1]))+ylab(expression(theta[2]))+ggtitle("Horseshoe Regression (E-Bayes)")
print(hs.2d.density)
```